---
title: "194CS - PA Road Network Analysis"
author: 'Brandon Wong II, '
date: "2026-03-05"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(igraph)
library(Matrix)
library(dplyr)
```


# Introduction

In this study, we are using the roadNet-PA dataset. It is a large road network from the Standord Network Analysis Project. It represents the road infrastructure of Pennsylvania as a graph. In this network, nodes represent road intersections, or endpoints, and edges represent road segments connecting those intersections. THe dataset contains about 1.09 million nodes and 1.5 million edges, making it a large-scale real-world network, commonly used for studying network structure, connectivity, and resilience. Because it models an actual road system, it is useful for us to analyze how transportation networks respond to random failures and disruptions. Furthermore, by analyzing cluseters presenting in the graph, we can indicate the transportation hubs and metropolitan areas that serves as critical intersections.

Objectives of the analysis.

The network we chose is a Pennsylvania Road Network from SNAP, Stanford Network Analysis Project. The network has 1088092 nodes and 1541898 edges and is undirected with a vertex name attribute. 

Our analysis has two objectives:

- Can we detect the Pennsylvania "Town" Structure within our Network

- Determine what keeps these "Towns" connected to each other 

# Methodology
Steps followed in data processing.
Tools and packages used (e.g., R, igraph).

### Data Processing // Subgraph Selection

To import our dataset, we first downloaded the .txt file from the SNAP website. Then we read in our .txt file using read.table which reads our .txt file in table format and creates a datafame from it. Now, we have our network edges in a dataframe from which we can just call graph_from_data_frame to convert our dataframe of edges into an actual graph object. 

In doing this, we are given a graph object with 1088092 nodes and 3083796 edges, which is exactly double the number of edges we should have in our network. What happened here, is that our .txt file represents an undirected edge by listing it twice, i.e A -> B and B -> A, meaning when we call our graph_from_data_frame function, it treats these as two separate edges between the two nodes and thus we get double the edges we should. To fix this, we just called the simplify function on our importted network, which gets rid of duplicate edges and self-loops, leaving us with the correct number of nodes and edges.

Now we have our network successfully imported, however, our network is too large to visualize or run community detection algorithms on, so we instead extracted two types of subgraphs from our whole network:

- *Hub-Centric Subgraphs* - These are subgraphs that are built iteratively using high-degree nodes and their neighbors within 15 steps. The steps for doing this are as follows:

  - 1. Find the degrees of all remaining nodes, sort in descending order and pull the index of the highest degree node as the start_node (if on Iteration 1, this would just use the whole network, simple_pa_road)

  - 2. Using the start node, build out the neighborhood and extract the node ID's using ego(), these are your subgraphs neighborhood nodes

  - 3. Create the actual subgraph using induced_subgraph() with the remaining network (or full network) and neighborhood nodes

  - 4. Plot the generated subgraph plot(sub1, ...)

  - 5. Find the remaining nodes by deleting the neighborhood nodes you used from the remaining network nodes (or whole network nodes if Iteration 1)

  - 6. Repeat

- *Randomly Sampled Subgraphs* - These are the subgraphs built iteratively using randomly sampled nodes and their neighbors within 15 steps. We use these subgraphs to see if our "town" structure exists outside of the major hubs. The steps for generating these subgraphs are the same as generating our Hub-Centric subgraphs, except here we do not need to find the highest degree start_node and instead we just sample 1 node from our remaining nodes. 

### Tools and Packages

We used R as our programming language to perform our analysis.

The packages we used were as follows:

- igraph: Used to interact with our network

- Matrix: Used to create the adjacency matrices

- dplyr: Used for data manipulation and cleaning


# Analysis

Description and visualization of subgraphs.
Metrics and community detection results.
Interpretation of adjacency matrix reorderings.

## Generating Subgraphs

Here, we generate subgraphs iteratively using the following steps:

- 1. Find the degrees of all remaining nodes, sort in descending order and pull the index of the highest degree node as the start_node (if on Iteration 1, this would just use the whole network, simple_pa_road)

- 2. Using the start node, build out the neighborhood and extract the node ID's using ego(), these are your subgraphs neighborhood nodes

- 3. Create the actual subgraph using induced_subgraph() with the remaining network (or full network) and neighborhood nodes

- 4. Plot the generated subgraph plot(sub1, ...)

- 5. Find the remaining nodes by deleting the neighborhood nodes you used from the remaining network nodes (or whole network nodes if Iteration 1)

- 6. Repeat

Considerations:

**Why generate subgraphs?**

- We generate subgraphs because our whole network is simply too large to run many of the summary statistics (diameter, betweenness/closeness centrality, average.path.length, etc) or community detection algorithms.

**Why generate iteratively?**

- 

**Why use the highest degree nodes to build the neighborhoods?**

- We use the highest degree nodes because in a road network, nodes (or intersections) with high degrees (high number of connecting roads) are likely major intersections where many roads converge to. Essentially what this does is find the city centers or hubs that we can use to see how surrounding neighborhoods are structured. 

**Why order = 15?**

- We chose an order level of 15 because choosing higher order levels will result in having similar problems to using the whole network, that is, it is computationally expensive. And for lower order levels, we get pretty small subgraphs which would give less meaningful information. 

# Results
Key findings from the analysis.
Visual and textual summaries.


# Conclusion
Summary of insights gained.
Possible future directions for further analysis.


# References

List of references and datasets used.

“SNAP: Network Datasets: Pennsylvania Road Network.” Stanford.edu, 2026, snap.stanford.edu/data/roadNet-PA.html. Accessed 6 Mar. 2026.

# Appendix

```{r}
# Margin helpers
def_marg <- c(5.1, 4.1, 4.1, 2.1)
no_marg  <- c(0, 0, 0, 0)

pa_road_edges <- read.table(
  "~/Desktop/194final_SNA/roadNet-PA.txt",
  comment.char = "#",
  col.names = c("from", "to")
)

pa_road_network <- graph_from_data_frame(pa_road_edges, directed=FALSE)

# Removes duplicate edges / self-loops
simple_pa_road <- simplify(pa_road_network)
```

## Subgraphs generations using high degree nodes
```{r, cache = TRUE}
# Subgraph 1
# 1. Select the first starting node (highest degree)
sub1_deg <- degree(simple_pa_road, mode="all")
sub1_sorted_deg <- sort(sub1_deg, decreasing = T, index.return = T)

start_node_1 <- sub1_sorted_deg$ix[1]

# 2. Extract the neighborhood
# we use ego() to get the IDs directly
sub1_nodes <- ego(simple_pa_road, order = 15, nodes = start_node_1, mode = "all")[[1]]

# 3. Create the actual subgraph object for plotting/analysis
sub1 <- induced_subgraph(simple_pa_road, sub1_nodes)

plot(sub1,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Hub 1"
)

# Remove these nodes from the main graph to create a "remaining" network
# We use the names/indices to delete them
remaining_network <- delete_vertices(simple_pa_road, sub1_nodes)


# Subgraph 2 
# Recalculate degrees on the REMAINING network to find the new "best" start node
sub2_deg <- degree(remaining_network)
sub2_sorted_deg <- sort(sub2_deg, decreasing = TRUE, index.return = TRUE)

# Select the new highest degree node from what's left
start_node_2 <- sub2_sorted_deg$ix[1]

# Extract the neighborhood from the remaining network
sub2_nodes <- ego(remaining_network, order = 15, nodes = start_node_2, mode = "all")[[1]]
sub2 <- induced_subgraph(remaining_network, sub2_nodes)

plot(sub2,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Hub 2"
)

remaining_network1 <- delete_vertices(remaining_network, sub2_nodes)

# Subgraph 3
sub3_deg <- degree(remaining_network1)
sub3_sorted_deg <- sort(sub3_deg, decreasing = TRUE, index.return = TRUE)

start_node_3 <- sub3_sorted_deg$ix[1]

sub3_nodes <- ego(remaining_network1, order = 15, nodes = start_node_3, mode = "all")[[1]]
sub3 <- induced_subgraph(remaining_network1, sub3_nodes)

plot(sub3,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Hub 3"
)

remaining_network2 <- delete_vertices(remaining_network1, sub3_nodes)

# Subgraph 4 
sub4_deg <- degree(remaining_network2)
sub4_sorted_deg <- sort(sub4_deg, decreasing = TRUE, index.return = TRUE)

# Select the new highest degree node from what's left
start_node_4 <- sub4_sorted_deg$ix[1]

sub4_nodes <- ego(remaining_network2, order = 15, nodes = start_node_4, mode = "all")[[1]]
sub4 <- induced_subgraph(remaining_network2, sub4_nodes)

plot(sub4,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Hub 4"
)

remaining_network3 <- delete_vertices(remaining_network2, sub4_nodes)

# Subgraph 5 
sub5_deg <- degree(remaining_network3)
sub5_sorted_deg <- sort(sub5_deg, decreasing = TRUE, index.return = TRUE)

start_node_5 <- sub5_sorted_deg$ix[1]

sub5_nodes <- ego(remaining_network3, order = 15, nodes = start_node_5, mode = "all")[[1]]
sub5 <- induced_subgraph(remaining_network3, sub5_nodes)

plot(sub5,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Hub 5"
)
```

### Shape interpretation for each subnetwork

**Focus on interpreting the first subgraph with one paragraph, and generalize the pattern for the rest with may be 1-3 sentences**

Subnetwork 1:

Subnetwork 2:
Subnetwork 3:
Subnetwork 4:
Subnetwork 5:

## Subgraphs Metrics
```{r, cache = TRUE}
my_subs <- list(sub1, sub2, sub3, sub4, sub5)
my_nodes <- c(start_node_1, start_node_2, start_node_3, start_node_4, start_node_5)
sub_names <- c("Hub 1", "Hub 2", "Hub 3", "Hub 4", "Hub 5")

results <- data.frame()

for(i in 1:length(my_subs)){
  g <- my_subs[[i]]
  
  # Community Detection
  cw <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Avg_Degree = mean(degree(g)),
    Transitivity = transitivity(g),
    Density = edge_density(g),
    Close_Centrality = centr_clo(g, mode="all", normalized=T)$centralization,
    Btwn_Centrality = centr_betw(g, directed=F, normalized=T)$centralization,
    Modularity_Walk = modularity(cw),
    Modularity_FG = modularity(cfg),
    Num_Communities_Walk = length(unique(membership(cw))),
    Num_Communities_FG = length(unique(membership(cfg))),
    Diameter = diameter(g) 
  )
  results <- rbind(results, row)
}

print(results)
```

### Mectrics Interpretation
**1-2 Paragraph**

## 3 Random subgraphs
```{r}
# Subgraph 1

sub_nodes1 <- ego(simple_pa_road, order = 15, nodes = sample(V(simple_pa_road), 1), mode = "all")[[1]]

# 3. Create the actual subgraph object for plotting/analysis
sub_1 <- induced_subgraph(simple_pa_road, sub_nodes1)

plot(sub_1,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Random Hub 1"
)

# Remove these nodes from the main graph to create a "remaining" network
# We use the names/indices to delete them
remaining_net <- delete_vertices(simple_pa_road, sub_nodes1)


# Subgraph 2 

sub_nodes2 <- ego(remaining_net, order = 15, nodes = sample(V(remaining_net), 1), mode = "all")[[1]]
sub_2 <- induced_subgraph(remaining_net, sub_nodes2)

plot(sub_2,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Random Hub 2"
)

remaining_net1 <- delete_vertices(remaining_net, sub_nodes2)

# Subgraph 3
sub_nodes3 <- ego(remaining_net1, order = 15, nodes = sample(V(remaining_net1), 1), mode = "all")[[1]]
sub_3 <- induced_subgraph(remaining_net1, sub_nodes3)

plot(sub_3,
     layout = layout_nicely,
     # === vertex properties 
     vertex.color = "yellow",
     vertex.frame.color = "black",
     vertex.shape = "square",
     vertex.size = 3,
     # === vertex label properties
     vertex.label = NA,
     # === edge properties
     edge.color = "black",
     edge.lty = 1,
     edge.width = 0.5,
     edge.arrow.size = 0.2,
     main = "Random Hub 3"
)
```

### Shape Interpretation

**(1-2 paragraphs) Briefly interpret all 3 subgraphs initialized with randomly sampled node, compare the structure of these subgraphs to subgraphs initialized with high degree node with 1-3 sentences**

## Random subgraphs metrics
```{r}
my_subs <- list(sub_1, sub_2, sub_3)
sub_names <- c("Subgraph 1", "Subgraph 2", "Subgraph 3")

random_results <- data.frame()

for(i in 1:length(my_subs)){
  g <- my_subs[[i]]
  
  # Community Detection
  cw <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Avg_Degree = mean(degree(g)),
    Transitivity = transitivity(g),
    Density = edge_density(g),
    Close_Centrality = centr_clo(g, mode="all", normalized=T)$centralization,
    Btwn_Centrality = centr_betw(g, directed=F, normalized=T)$centralization,
    Modularity_Walk = modularity(cw),
    Modularity_FG = modularity(cfg),
    Num_Communities_Walk = length(unique(membership(cw))),
    Num_Communities_FG = length(unique(membership(cfg))),
    Diameter = diameter(g) 
  )
  random_results <- rbind(random_results, row)
}

print(random_results)
```

### Mectrics Intepretation
**Interpret these random node subgraphs mectrics for 1 paragraph**

## Community Detection (subnetwork 1)
```{r, cache = TRUE}
set.seed(932085) 

cw <- cluster_walktrap(sub1)
cfg <- cluster_fast_greedy(sub1)


# Color palette
all_colors <- colors()
vibrant_colors <- all_colors[!grepl("gray|grey", all_colors)]
unique_palette <- sample(vibrant_colors)


# Plotting Walktrap
V(sub1)$community <- cw$membership
num_comms <- length(unique(V(sub1)$community))
com_colors <- unique_palette[1:num_comms]
V(sub1)$color <- com_colors[V(sub1)$community]

par(mar = no_marg)
plot(sub1,
     vertex.size = 4,
     vertex.label = NA,
     vertex.color = V(sub1)$color,
     vertex.frame.color = "gray40",
     edge.color = "black",
     edge.width = 0.5,
     edge.curved = 0.3,
     mark.groups = communities(cw),
     mark.col = "grey95",
     mark.border = "red",
     main = "Walktrap Communities (colored hulls)")


# Plotting Fast and Greedy
V(sub1)$community <- cfg$membership
num_comms <- length(unique(V(sub1)$community))
com_colors <- unique_palette[1:num_comms]
V(sub1)$color <- com_colors[V(sub1)$community]

par(mar = no_marg)
plot(sub1,
     vertex.size = 4,
     vertex.label = NA,
     vertex.color = V(sub1)$color,
     vertex.frame.color = "gray40",
     edge.color = "black",
     edge.width = 0.5,
     edge.curved = 0.3,
     mark.groups = communities(cw),
     mark.col = "grey95",
     mark.border = "red",
     main = "F&G Communities (colored hulls)")
```

### Walktrap
**Explain the walktrap algorithm**

### Fast Greedy
**Explain the fast greedy algorithm**

### Subgraph Interpretation
**Interpret the plots on subnetwork1 with different community detection method**


## Adjacency Matrix Reodering (subnetwork 1)
```{r}
# Original (unordered) adjacency matrix
A <- as_adjacency_matrix(sub1, sparse = TRUE)
image(A, main = "Adjacency Matrix - original node order", col.region = c("red"),
      xlab = "Node", ylab = "Node")

# Reorder by Walktrap community membership
ord <- order(cw$membership)
A_reordered <- A[ord, ord]

image(A_reordered,
      main = "Adjacency Matrix — reordered by community (Walktrap)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")

# Reorder by Fast Greedy community membership
ord <- order(cfg$membership)
A_reordered1 <- A[ord, ord]

image(A_reordered1,
      main = "Adjacency Matrix — reordered by community (Fast Greedy)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")
```

### Adjacency Matrix Intepretation

**Intepret adjacency matrix for one paragraph**

## Resilence Analysis

### Objectives
**Objective of testing the resilience**


```{r}
# This chunk gives us the articulation points for each subnetwork, and the betweenness centrality of all articulation points within each subnetwork. And I store the top 5 highest centrality articulation nodes to be removed


artic_points1 <- articulation_points(sub1)
cat("Number of articulation points for hub 1:", length(artic_points1), '\n')
betweenness_sub1 <- betweenness(sub1, 
                                v = artic_points1,
                                directed = FALSE)
high_between_artic_node1 <- sort(betweenness_sub1, decreasing = T, index.return = T)$ix[1:5]
artic_points2 <- articulation_points(sub2)
cat("Number of articulation points for hub 2:", length(artic_points2), '\n')
betweenness_sub2 <- betweenness(sub2, 
                                v = artic_points2,
                                directed = FALSE)
high_between_artic_node2 <- sort(betweenness_sub2, decreasing = T, index.return = T)$ix[1:5]
artic_points3 <- articulation_points(sub3)
cat("Number of articulation points for hub 3:", length(artic_points3), '\n')
betweenness_sub3 <- betweenness(sub3, 
                                v = artic_points3,
                                directed = FALSE)
high_between_artic_node3 <- sort(betweenness_sub3, decreasing = T, index.return = T)$ix[1:5]
artic_points4 <- articulation_points(sub4)
cat("Number of articulation points for hub 4:", length(artic_points4), '\n')
betweenness_sub4 <- betweenness(sub4, 
                                v = artic_points4,
                                directed = FALSE)
high_between_artic_node4 <- sort(betweenness_sub4, decreasing = T, index.return = T)$ix[1:5]
artic_points5 <- articulation_points(sub5)
cat("Number of articulation points for hub 5:", length(artic_points5), '\n')
betweenness_sub5 <- betweenness(sub5, 
                                v = artic_points5,
                                directed = FALSE)
high_between_artic_node5 <- sort(betweenness_sub5, decreasing = T, index.return = T)$ix[1:5]
```
**Explain why remove high betweenness articulation point**


```{r eval = F, echo = F}
# These prints the 5 nodes for each subnetwork
high_between_artic_node1
high_between_artic_node2
high_between_artic_node3
high_between_artic_node4
high_between_artic_node5
```
**Explain the removal for 1-2 sentences**
remove the 5 nodes in each network
```{r}
removal_sub1 <- delete_vertices(sub1, high_between_artic_node1)
removal_sub2 <- delete_vertices(sub2, high_between_artic_node1)
removal_sub3 <- delete_vertices(sub3, high_between_artic_node1)
removal_sub4 <- delete_vertices(sub4, high_between_artic_node1)
removal_sub5 <- delete_vertices(sub5, high_between_artic_node1)
```


```{r}
removal_results <- data.frame()
removal_subs <- list(sub1, removal_sub1, sub2, removal_sub2, sub3, removal_sub3, sub4, removal_sub4, sub5, removal_sub5)
sub_names <- c("Hub 1", "Removal Hub 1", "Hub 2","Removal Hub 2", "Hub 3","Removal Hub 3", "Hub 4", "Removal Hub 4", "Hub 5", "Removal Hub 5")

for(i in 1:length(removal_subs)){
  g <- removal_subs[[i]]
  
  # Community Detection
  cw <- cluster_walktrap(g)
  
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Num_Comp = components(g)$no,
    Num_Disconnected_Pairs = distance_table(g,  directed = FALSE)$unconnected,
    Largest_Comp = max(components(g)$csize),
    Transitivity = round(transitivity(g),4),
    Diameter = diameter(g),
    Btwn_Centrality = round(centr_betw(g, directed = F)$centralization,4),
    Close_Centrality = round(centr_clo(g)$centralization,4),
    Modularity_Walk = round(modularity(cw),4),
    Num_Communities_Walk = length(unique(membership(cw)))
  )
  removal_results <- rbind(removal_results, row)
}

removal_results$Pctg_Disconnected_Pairs <- round(removal_results$Num_Disconnected_Pairs /
          (removal_results$Nodes * (removal_results$Nodes - 1) / 2) * 100, 4)
```


```{r}
print(removal_results)
# print(results)
```

### Mectrics Comparison and Interpretation

**One paragraph to summarize the differences**
result: the number of connected components stay the same for sub3 and sub5, it is likely that these two networks is more like a town that does not need highway or bridge connections between regions/districts. For removal_sub1, 2, 4 which has `NaN` closeness centrality centralization value indicates that there are isolated nodes with no connections. Number of communities drop after removal of high betweenness articulation node

Those top 5 articulation points were major highway junctions or interchange nodes that connected otherwise distinct geographic regions
Their high betweenness means most cross-region traffic flowed through them
Community detection interpreted these bridge nodes as boundaries, creating more communities
After removal, the algorithm no longer sees those artificial boundaries, so neighboring regions consolidate



