---
title: "194CS - PA Road Network Analysis"
author: "Brandon Wong II, "
date: "2026-03-05"
output: pdf_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
library(Matrix)
library(dplyr)
library(kableExtra)
```


```{r, , include=FALSE, echo = FALSE}
# Margin helpers
def_marg <- c(5.1, 4.1, 4.1, 2.1)
no_marg  <- c(0, 0, 0, 0)

pa_road_edges <- read.table(
  "roadNet-PA.txt",
  comment.char = "#",
  col.names = c("from", "to")
)

pa_road_network <- graph_from_data_frame(pa_road_edges, directed=FALSE)

# Removes duplicate edges / self-loops
simple_pa_road <- igraph::simplify(pa_road_network)
```

# Introduction

The roadNet-PA dataset found on Stanford's Network Analysis Project, or SNAP for short, is the subject of our analysis. This dataset represents the Pennsylvania road infrastructure as an undirected graph network where **nodes** (or vertices) are intersections or endpoints (dead-ends / cul-de-sacs) and **edges** are the roads that connect them. The network has rough 1.9 million nodes and 1.5 million edges which makes it a relatively comprehensive model of a real-world transportation system. 

Our analysis has two primary objectives:

- "Town" Detection: Can we find distinct Functional Neighborhoods or "Towns" within major hubs using community detection algorithms?

- "Town" Connectivity & Resilience: How does the removal of high-betweenness articulation points impact the structural resilience of these "Towns"? In other words, is Pennsylvania's road infrastructure susceptible to major intersection closures or does it offer alternate paths for traversing the network?

# Methodology

### Data Processing

To import our dataset, we first downloaded the .txt file (roadNet-PA.txt) from the SNAP website. Then, we imported the data using `read.table` which reads our .txt file in table format and creates a datafame from it. Now, we have our network edges in a dataframe from which we can just call `graph_from_data_frame` to convert our dataframe of edges into an actual graph object. 

After doing this, we have a graph object with 1088092 nodes and 3083796 edges, which is exactly double the number of edges we should have in our network. What happened here, is that our .txt file represents an undirected edge by listing it twice, i.e A -> B and B -> A, meaning when we call our graph_from_data_frame function, it treats these as two separate edges between the two nodes and thus we get double the edges we should. To fix this, we called the simplify function on our imported network, which removes duplicate edges and self-loops, leaving us with the correct number of nodes and edges in our network.

### Subgraph Generation

Due to how computationally expensive it is to visualize and analyze 1 million+ nodes, we needed to come up with a way to generate subgraphs. The sampling method we employed is as follows:

1. Find the degrees of all remaining nodes, sort in descending order and pull the index of the highest degree node as the start node (if on Iteration 1, use the whole network: simple_pa_road)

2. Using found start node, build out the neighborhood at an order of 15 steps and extract the node ID's using ego(), these are your "urban centers" nodes that will be used to generate the subgraph.  

3. Create the actual subgraph using `induced_subgraph()` with the remaining network (or full network) and urban center nodes

4. Plot the generated subgraph `plot(sub1, ...)`

5. Find the remaining nodes by deleting the neighborhood nodes you used from the remaining network nodes (or whole network nodes if Iteration 1)

6. Repeat

The main idea of this subgraph generation method is to mimic selecting urban centers and their surrounding areas. By iteratively choosing the highest-degree node of the network as the start node we build out neighborhoods to an order of 15 steps which effectively gives us subgraphs representing 5 different urban centers of PA's road network. 

### Community Detection Algorithms

**Walktrap** is a community detection algorithm that implement the idea of random walks to identify densely connected groups within a network. The intuition is that a random walker traversing a network tends to get trapped within a dense community, nodes being visited frequently are likely to be in the same community. The algorithm computes the distance between nodes based on short random walks (by default 4 steps), then uses this distance to perform hierarchical agglomerative clustering, iteratively merging the most similar nodes or communities until the entire graph collapses into one cluster. This maximize the modularity at some point in this merging hierarchy. The runtime complexity is $O(mn^2)$ in the worst case (where m is the number of edges and n is the number of nodes), making it practical for moderately sized networks. One drawback is that the computational cost is higher compared to fast greedy and louvain.

**Fast Greedy** is a community detection algorithm that greedily optimize modularity through a agglomerative (bottom-up) hierarchical clustering process. It starts by treating each node as its own separate community, then repeatedly merges pairs of communities that yield the largest increase in modularity at each step, modularity being a measure of how much more densely connected nodes are within communities compared to that in a random graph. The merging process of a community continues until further partition does not yield increases in modulariry. The runtime complexity is to $O(m \log n)$, making it scalable to large sparse networks like social or citation graphs. The process continues until all nodes belong to a single community, producing a full dendrogram, and the partition with the highest modularity score during the merging process is returned as the final community structure. One drawback is that it might overlook small communities.

### Choosing optimal number of removed High-Betweenness Articulation Points



### Tools and Packages

We used R and RStudio as our programming language and platform to perform our analysis.

The packages we used were as follows:

- igraph: Used for network manipulation, community detection, and metric calculations

- Matrix: Used to create the adjacency matrices

- dplyr: Used for data manipulation and cleaning

- kableExtra: Used to create better tables


# Analysis

### Description and Visualization of Subgraphs

To visualize our subgraphs, we used the igraph plot function with the layout function of layout_nicely. The Layout_nicely function chooses automatically the most appropriate layout algorithm for the graph. The reason for only choosing this layout, is that other layout algorithms are unideal for visualizing road networks, for example, layout_on_grid or layout_in_circle plot the nodes as described, which for a road network, doesn't make sense for intersections to be in a circle or on a uniform grid as this is not how a road network would look in reality. As such we use layout_nicely to plot our network since it automatically selects the most optimal layout for our subnetworks. 

```{r hub-generation, cache = TRUE, echo = FALSE}
set.seed(2249282)
current_network <- simple_pa_road
subgraph_list_hub <- list()
num_subgraphs <- 5

par(mfrow = c(2, 3),mar = c(0, 1, 1, 1))

for (i in 1:num_subgraphs) {
  
  # Calculate degrees and find the top node
  deg <- degree(current_network, mode = "all")
  sorted_deg <- sort(deg, decreasing = TRUE, index.return = TRUE)
  start_node <- sorted_deg$ix[1]
  
  # Extract the neighborhood (order 15)
  nodes_to_extract <- ego(current_network, order = 15, nodes = start_node, mode = "all")[[1]]
  
  # Create the subgraph
  sub <- induced_subgraph(current_network, nodes_to_extract)
  
  # Store it in our list in case we want to analyze it later
  subgraph_list_hub[[i]] <- sub
  
  # Plot the subgraph immediately
  plot(sub,
       layout = layout_nicely,
       main = paste("Hub", i),
       vertex.color = "yellow",
       vertex.frame.color = "black",
       vertex.shape = "square",
       vertex.size = 3,
       vertex.label = NA,
       edge.color = "black",
       edge.lty = 1,
       edge.width = 0.5,
       edge.arrow.size = 0.2)
  
  # Update the network by removing the extracted nodes for the next iteration
  current_network <- delete_vertices(current_network, nodes_to_extract)
}
```

Above we can see the generated subgraphs. These subgraphs offer us a look into the topological structure of Pennsylvania's Road Network. Since these were built-out using the ego() function and high-degree nodes as the origin, each of these subgraphs serve to represent a major region in Pennsylvania. 

Hubs 1 and 2 we can see long-sweeping edges that bypass local clusters of intersections. This suggests that our high-degree nodes used to generate these subgraphs are located near major highways, where these highways connect the local intersection clusters.

Hubs 3 and 5 are characterized by their dense, grid-like structure. These are likely major metropolitan areas of Pennsylvania where the grid-like structure represents the urban planning, where there is multiple routes existing between 2 points. 

Hub 4 is the most unique out of the 5 subnetworks. We can see a clear central core with "arms" that branch out from it. These "arms" are not connected to each other and only connected through the central core, meaning if you removed the node connecting the core to this arm, then this arm would be completely isolated from the network. This shows to us how central town or city acts as the connection to smaller towns or rural roads.

Overall, these hubs reveal various different topological structures of Pennsylvanis Road Network. Revealing why we chose to build out networks from high-degree nodes. 

### Network Metrics

```{r hub-results , cache = TRUE, echo = FALSE}
results_hub <- data.frame()
sub_names <- c("Hub 1", "Hub 2", "Hub 3", "Hub 4", "Hub 5")
for(i in 1:length(subgraph_list_hub)){
  g <- subgraph_list_hub[[i]]
  g <- simplify(g)
  
  # Community Detection
  cw <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Avg_Degree = mean(degree(g)),
    Transitivity = transitivity(g),
    Density = edge_density(g),
    Close_Centrality = centr_clo(g, mode="all", normalized=T)$centralization,
    Btwn_Centrality = centr_betw(g, directed=F, normalized=T)$centralization,
    Modularity_Walk = modularity(cw),
    Modularity_FG = modularity(cfg),
    Num_Communities_Walk = length(unique(membership(cw))),
    Num_Communities_FG = length(unique(membership(cfg))),
    Diameter = diameter(g) 
  )
  results_hub <- rbind(results_hub, row)
}

print(results_hub)
```



**Edge**: road segment directly connecting two nodes/ intersections. Like the road connecting node A and node B where it typical records as two edges between two nodes. But, we just directly record one edge between them, applying simplify function.

**Adv_degree**: means the average degree of each note connection. In our dataset, it means the average number of roads connecting at each intersection in PA. For example, in the dataset, all hubs have average degree between **2.53 and 3.28**, meaning each Pennsylvania intersection typically connects to 2-3 other intersections.

Hub3 has the largest average degree of 3.28, which indicates on average, each node in Hub3 has approximately 3 roads segments connecting to other intersections. suggesting more crossroad-style intersections

Hub 4 has the lowest (2.53) suggesting more dead ends and T-junctions.

**Transitivity**:
**In road context:** If intersection A connects to B, and B connects to C, how likely does C also connect back to A forming a road loop or city block.

**Interpretation:**
Looking through hub1 to hub5, we find that all hubs show very low transitivity between **0.045 and 0.071**, meaning only 4-7% of possible road triangles actually exist. Hub 4 has the highest (0.071) suggesting more city-block style loops, while Hub 5 has the lowest (0.045) suggesting mostly straight linear roads.

**Density**:
*In road context:* Out of every possible road that could connect any two intersections, what fraction actually exists.

*Interpretation:* > All hubs have extremely low density between **0.0016 and 0.0049**, meaning less than 0.5% of all possible road connections exist. Hub 3 is densest (0.0049) despite being the smallest, suggesting a compact well-connected neighborhood. Hub 2 is least dense (0.0016) despite having the most nodes — suggesting a large spread-out suburban area where intersections are far apart.

**Close_Centrality**: How few road segments an intersection needs to travel through to reach any other intersection in the neighborhood. Higher closeness means better **traffic flow and accessibility** throughout the area.
- *Interpretation:* Take hub3 as an example.It has the lowest closeness (0.050). It means drivers need **more turns and road segments** to reach distant intersections.

**Btwn_Centrality**: How many driver routes must pass through a given intersection — identifying bottleneck intersections critical to traffic flow. High betweenness hubs are most **vulnerable to traffic disruption** if key intersections are closed or damaged.
*Interpretation:* Hub 3 has the lowest betweenness (0.188) suggesting more **alternative routes exist** — drivers have more flexibility to avoid any single intersection.

**Modularity & Communities**: How strongly Pennsylvania roads cluster into distinct local neighborhoods with few roads connecting between them.

*Interpretation:* All hubs show extremely high modularity between **0.80 and 0.90**. This confirms Pennsylvania roads naturally form **tight geographic clusters** with few cross-neighborhood connections.

Fast Greedy consistently finds fewer communities (16-31) than Walktrap (27-112), but both agree on strong community structure. Hub 2 has the most communities (112 by Walktrap) reflecting its large size and many disconnected sub-neighborhoods, while Hub 3 has the fewest (27) reflecting its compact connected structure.

### Community Detection Results 

```{r community-detection-visualization, cache = TRUE, echo = FALSE}
all_colors <- colors()
vibrant_colors <- all_colors[!grepl("gray|grey", all_colors)]
unique_palette <- sample(vibrant_colors)

par(mfrow = c(2,2), mar = c(0, 1, 1, 0))
# Loop through each subgraph in your list
for(i in c(2,3)) {
  g <- subgraph_list_hub[[i]]
  
  cw  <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  
  # Walktrap
  num_comms_cw <- length(unique(membership(cw)))
  v_colors_cw <- unique_palette[1:num_comms_cw]
  
  plot(cw, g,
       layout = layout_nicely,
       vertex.size = 4,
       vertex.label = NA,
       vertex.color = v_colors_cw[membership(cw)],
       vertex.frame.color = "gray40",
       edge.color = "black",
       edge.width = 0.5,
       edge.curved = 0.3,
       col = "grey95",      # hull color
       border = "red",      # hull border
       main = paste("Hub", i, ": Walktrap"))

  # Fast & Greedy
  num_comms_cfg <- length(unique(membership(cfg)))
  v_colors_cfg <- unique_palette[1:num_comms_cfg]
  
  plot(cfg, g,
       layout = layout_nicely,
       vertex.size = 4,
       vertex.label = NA,
       vertex.color = v_colors_cfg[membership(cfg)],
       vertex.frame.color = "gray40",
       edge.color = "black",
       edge.width = 0.5,
       edge.curved = 0.3,
       col = "grey95",
       border = "blue",     # changed border color to distinguish
       main = paste("Hub", i, ": F&G"))
}
```

Walktrap produces more smaller and fragmented communities, which is visible from the small isolated colored clusters near the red boundaries, reflecting its sensitivity to local structure through short random walks. The small communities detected are likely to be neighborhoods or blocks that are closely connected by local streets rather than major roadways connecting different part of the town.

In contrast, Fast Greedy produces fewer, larger, and more cohesive communities, which is consistent with its global modularity optimization that tends to miss small communities and merge those communities into larger ones. The larger communities detected are more widespread than those in walktrap and thus likely to be larger districts or regions of the town/city that includes multiple neighborhoods or blocks, and these communities are likely connected by major roadway in the town with high betweenness centrality.

### Adjacency Matrix Reordering

```{r adj-mat-reorder, cache = TRUE, echo = FALSE}
cw <- cluster_walktrap(subgraph_list_hub[[1]])
cfg <- cluster_fast_greedy(subgraph_list_hub[[1]])

# Original (unordered) adjacency matrix
A <- as_adjacency_matrix(subgraph_list_hub[[1]], sparse = TRUE)
image(A, main = "Original Node Order", col.region = c("red"),
      xlab = "Node", ylab = "Node")

# Reorder by Walktrap community membership
ord <- order(cw$membership)
A_reordered <- A[ord, ord]

image(A_reordered,
      main = "Reordered by Community (Walktrap)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")

# Reorder by Fast Greedy community membership
ord <- order(cfg$membership)
A_reordered1 <- A[ord, ord]

image(A_reordered1,
      main = "Reordered by Community (Fast Greedy)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")
```

### Resilience Analysis

```{r, echo = FALSE}
sub1 <- subgraph_list_hub[[1]]
sub2 <- subgraph_list_hub[[2]]
sub3 <- subgraph_list_hub[[3]]
sub4 <- subgraph_list_hub[[4]]
sub5 <- subgraph_list_hub[[5]]

artic_points1 <- articulation_points(sub1)
# cat("Number of articulation points for hub 1:", length(artic_points1), '\n')
betweenness_sub1 <- betweenness(sub1, v = artic_points1,directed = FALSE)
high_between_artic_node1 <- sort(betweenness_sub1, decreasing = T, index.return = T)$ix[1:5]
artic_points2 <- articulation_points(sub2)
# cat("Number of articulation points for hub 2:", length(artic_points2), '\n')
betweenness_sub2 <- betweenness(sub2, v = artic_points2, directed = FALSE)
high_between_artic_node2 <- sort(betweenness_sub2, decreasing = T, index.return = T)$ix[1:5]
artic_points3 <- articulation_points(sub3)
# cat("Number of articulation points for hub 3:", length(artic_points3), '\n')
betweenness_sub3 <- betweenness(sub3, v = artic_points3, directed = FALSE)
high_between_artic_node3 <- sort(betweenness_sub3, decreasing = T, index.return = T)$ix[1:5]
artic_points4 <- articulation_points(sub4)
# cat("Number of articulation points for hub 4:", length(artic_points4), '\n')
betweenness_sub4 <- betweenness(sub4, v = artic_points4, directed = FALSE)
high_between_artic_node4 <- sort(betweenness_sub4, decreasing = T, index.return = T)$ix[1:5]
artic_points5 <- articulation_points(sub5)
# cat("Number of articulation points for hub 5:", length(artic_points5), '\n')
betweenness_sub5 <- betweenness(sub5, v = artic_points5, directed = FALSE)
high_between_artic_node5 <- sort(betweenness_sub5, decreasing = T, index.return = T)$ix[1:5]

# table includes name of hub and number of arrticulation point
artic_df <- tibble(Name=c("Hub 1", "Hub 2", "Hub 3", "Hub 4", "Hub 5"), `Articulation Points`=as.numeric(c(length(artic_points1), length(artic_points2), length(artic_points3), length(artic_points4), length(artic_points5))))
kable(t(artic_df), col.names = NULL, caption = "Number of Articulation points for each hub")
```

Removing high betweenness articulation node is particularly impactful because these nodes simultaneously serve as connectors that removal disconnects the network and as major traffic bottlenecks through which the most shortest paths flow. Their failure causes both structural fragmentation and disruption to the overall reachability of the road network because people need to find another route to commute from place to place with higher commute cost, such as longer commute time and increase in gas usage.

```{r, echo = FALSE}
# Removing High-Betweenness Articulation Points
removal_sub1 <- delete_vertices(sub1, high_between_artic_node1)
removal_sub2 <- delete_vertices(sub2, high_between_artic_node1)
removal_sub3 <- delete_vertices(sub3, high_between_artic_node1)
removal_sub4 <- delete_vertices(sub4, high_between_artic_node1)
removal_sub5 <- delete_vertices(sub5, high_between_artic_node1)

removal_results <- data.frame()
removal_subs <- list(sub1, removal_sub1, sub2, removal_sub2, sub3, removal_sub3, sub4, removal_sub4, sub5, removal_sub5)
sub_names <- c("Hub 1", "Removal Hub 1", "Hub 2","Removal Hub 2", "Hub 3","Removal Hub 3", "Hub 4", "Removal Hub 4", "Hub 5", "Removal Hub 5")

for(i in 1:length(removal_subs)){
  g <- removal_subs[[i]]
  
  # Community Detection
  cw <- cluster_walktrap(g)
  
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Num_Comp = components(g)$no,
    Num_Disconnected_Pairs = distance_table(g,  directed = FALSE)$unconnected,
    Largest_Comp = max(components(g)$csize),
    Transitivity = round(transitivity(g),4),
    Diameter = diameter(g),
    Btwn_Centrality = round(centr_betw(g, directed = F)$centralization,4),
    Close_Centrality = round(centr_clo(g)$centralization,4),
    Modularity_Walk = round(modularity(cw),4),
    Num_Communities_Walk = length(unique(membership(cw)))
  )
  removal_results <- rbind(removal_results, row)
}

removal_results$Pctg_Disconnected_Pairs <- round(removal_results$Num_Disconnected_Pairs /
          (removal_results$Nodes * (removal_results$Nodes - 1) / 2) * 100, 4)

t(removal_results) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped") %>%
  column_spec(c(2,3), background = "#FFE4E1") %>%  # Hub 1
  column_spec(c(4,5), background = "#E1F5E1") %>%  # Hub 2
  column_spec(c(6,7), background = "#E1E8FF") %>%  # Hub 3
  column_spec(c(8,9), background = "#FFF8E1") %>%  # Hub 4
  column_spec(c(10,11), background = "#F3E1FF")
```

the number of connected components stay the same for sub3 and sub5, it is likely that these two networks is more like a town that does not need highway or bridge connections between regions/districts. For removal_sub1, 2, 4 which has `NaN` closeness centrality centralization value indicates that there are isolated nodes with no connections. Number of communities drop after removal of high betweenness articulation node

Those top 5 articulation points were major highway junctions or interchange nodes that connected otherwise distinct geographic regions Their high betweenness means most cross-region traffic flowed through them Community detection interpreted these bridge nodes as boundaries, creating more communities After removal, the algorithm no longer sees those artificial boundaries, so neighboring regions consolidate

# Results
Key findings from the analysis.
Visual and textual summaries.


# Conclusion
Summary of insights gained.
Possible future directions for further analysis.


# References

List of references and datasets used.

“SNAP: Network Datasets: Pennsylvania Road Network.” Stanford.edu, 2026, snap.stanford.edu/data/roadNet-PA.html. Accessed 6 Mar. 2026.

# Appendix

```{r data-processing}
# Data Processing 

# Margin helpers
def_marg <- c(5.1, 4.1, 4.1, 2.1)
no_marg  <- c(0, 0, 0, 0)

pa_road_edges <- read.table(
  "roadNet-PA.txt",
  comment.char = "#",
  col.names = c("from", "to")
)

pa_road_network <- graph_from_data_frame(pa_road_edges, directed=FALSE)

# Removes duplicate edges / self-loops
simple_pa_road <- simplify(pa_road_network)
```

       

