---
title: "194CS - PA Road Network Analysis"
author: "Brandon Wong II, "
date: "2026-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
library(Matrix)
library(dplyr)
set.seed(2048338)
```


# Introduction

The roadNet-PA dataset found on Stanford's Network Analysis Project, or SNAP for short, is the subject of our analysis. This dataset represents the Pennsylvania road infrastructure as an undirected graph network where **nodes** (or vertices) are intersections or endpoints (dead-ends / cul-de-sacs) and **edges** are the roads that connect them. The network has rough 1.9 million nodes and 1.5 million edges which makes it a relatively comprehensive model of a real-world transportation system. 

Our analysis has two primary objectives:

- "Town" Detection: Can we find distinct Functional Neighborhoods or "Towns" within major hubs using community detection algorithms and how does this "town" structure differ between major hubs (urban centers) and randomly sampled hubs (rural regions)?

- "Town" Connectivity & Resilience: How does the removal of high-betweenness articulation points impact the structural resilience of these "Towns"? In other words, is Pennsylvania's road infrastructure susceptible to major intersection closures or does it offer alternate paths for traversing the network?

# Methodology

### Data Processing

To import our dataset, we first downloaded the .txt file (roadNet-PA.txt) from the SNAP website. Then, we imported the data using `read.table` which reads our .txt file in table format and creates a datafame from it. Now, we have our network edges in a dataframe from which we can just call `graph_from_data_frame` to convert our dataframe of edges into an actual graph object. 

After doing this, we have a graph object with 1088092 nodes and 3083796 edges, which is exactly double the number of edges we should have in our network. What happened here, is that our .txt file represents an undirected edge by listing it twice, i.e A -> B and B -> A, meaning when we call our graph_from_data_frame function, it treats these as two separate edges between the two nodes and thus we get double the edges we should. To fix this, we called the simplify function on our imported network, which removes duplicate edges and self-loops, leaving us with the correct number of nodes and edges in our network.

### Subgraph Generation

Due to how computationally expensive it is to visualize and analyze 1 million+ nodes, we made use of two similar subgraph generation strategies:

**Hub-Centric Subgraphs** - The main idea of this subgraph generation is to mimic selecting urban centers and their surrounding areas. By iteratively choosing the highest-degree node of the network as the start node we build out neighborhoods to an order of 15 steps which effectively gives us subgraphs representing 5 different urban centers of PA's road network. 

The steps we used to do this are as follows:

1. Find the degrees of all remaining nodes, sort in descending order and pull the index of the highest degree node as the start node (if on Iteration 1, use the whole network: simple_pa_road)

2. Using found start node, build out the neighborhood at an order of 15 steps and extract the node ID's using ego(), these are your "urban centers" nodes that will be used to generate the subgraph.  

3. Create the actual subgraph using induced_subgraph() with the remaining network (or full network) and urban center nodes

4. Plot the generated subgraph plot(sub1, ...)

5. Find the remaining nodes by deleting the neighborhood nodes you used from the remaining network nodes (or whole network nodes if Iteration 1)

6. Repeat

**Randomly Sampled Subgraphs** - These subgraphs were generated in a similar manner to the Hub-Centric Subgraphs, where the only difference is that instead of selecting the highest degree node as the starting point, we randomly sample 1 node from all / remaining nodes as the start node. The idea behind generating these subgraphs is to see if our "town" structure exists outside of the major hubs. 

### Tools and Packages

We used R and RStudio as our programming language and platform to perform our analysis.

The packages we used were as follows:

- igraph: Used for network manipulation, community detection, and metric calculations

- Matrix: Used to create the adjacency matrices

- dplyr: Used for data manipulation and cleaning


# Analysis

Description and visualization of subgraphs.
Metrics and community detection results.
Interpretation of adjacency matrix reorderings.

### Description and Visualization of Subgraphs

To visualize our subgraphs, we used the igraph plot function with the layout function of layout_nicely. The Layout_nicely function chooses automatically the most appropriate layout algorithm for the graph. The reason for only choosing this layout, is that other layout algorithms are unideal for visualizing road networks, for example, layout_on_grid or layout_in_circle plot the nodes as described, which for a road network, doesn't make sense for intersections to be in a circle or on a uniform grid as this is not how a road network would look in reality. As such we use layout_nicely to plot our network since it automatically selects the most optimal layout for our subnetworks. 

Comparing the Hub-Centric Subgraphs to the Randomly Sampled Subgraphs, we can see that our Hub-Centric subgraphs tend to have much more nodes or intersections than the Randomly Sampled Subgraphs. This reveals that our methods for generating these subgraphs did what we want them to, that is, 

have areas with much denser collections of nodes or intersections which reveals to us the neighborhoods or town structure of these urban center whereas for our Randomly Sampled Subgraphs, they have much more sparse groups of intersections 


### Network Metrics



### Community Detection Results 



### Adjacency Matrix Reordering



# Results
Key findings from the analysis.
Visual and textual summaries.


# Conclusion
Summary of insights gained.
Possible future directions for further analysis.


# References

List of references and datasets used.

“SNAP: Network Datasets: Pennsylvania Road Network.” Stanford.edu, 2026, snap.stanford.edu/data/roadNet-PA.html. Accessed 6 Mar. 2026.

# Appendix

```{r data-processing}
# Data Processing 

# Margin helpers
def_marg <- c(5.1, 4.1, 4.1, 2.1)
no_marg  <- c(0, 0, 0, 0)

pa_road_edges <- read.table(
  "roadNet-PA.txt",
  comment.char = "#",
  col.names = c("from", "to")
)

pa_road_network <- graph_from_data_frame(pa_road_edges, directed=FALSE)

# Removes duplicate edges / self-loops
simple_pa_road <- simplify(pa_road_network)
```


```{r hub-generation, cache = TRUE}
current_network <- simple_pa_road
subgraph_list_hub <- list()
num_subgraphs <- 5

for (i in 1:num_subgraphs) {
  
  # Calculate degrees and find the top node
  deg <- degree(current_network, mode = "all")
  sorted_deg <- sort(deg, decreasing = TRUE, index.return = TRUE)
  start_node <- sorted_deg$ix[1]
  
  # Extract the neighborhood (order 15)
  nodes_to_extract <- ego(current_network, order = 15, nodes = start_node, mode = "all")[[1]]
  
  # Create the subgraph
  sub <- induced_subgraph(current_network, nodes_to_extract)
  
  # Store it in our list in case we want to analyze it later
  subgraph_list_hub[[i]] <- sub
  
  # Plot the subgraph immediately
  plot(sub,
       layout = layout_nicely,
       main = paste("Hub", i),
       vertex.color = "yellow",
       vertex.frame.color = "black",
       vertex.shape = "square",
       vertex.size = 3,
       vertex.label = NA,
       edge.color = "black",
       edge.lty = 1,
       edge.width = 0.5,
       edge.arrow.size = 0.2)
  
  # Update the network by removing the extracted nodes for the next iteration
  current_network <- delete_vertices(current_network, nodes_to_extract)
}
```


```{r hub-results , cache = TRUE}
results_hub <- data.frame()

for(i in 1:length(subgraph_list_hub)){
  g <- subgraph_list_hub[[i]]
  
  # Community Detection
  cw <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  # Build row
  row <- data.frame(
    Name = sub_names[i],
    Nodes = vcount(g),
    Edges = ecount(g),
    Avg_Degree = mean(degree(g)),
    Transitivity = transitivity(g),
    Density = edge_density(g),
    Close_Centrality = centr_clo(g, mode="all", normalized=T)$centralization,
    Btwn_Centrality = centr_betw(g, directed=F, normalized=T)$centralization,
    Modularity_Walk = modularity(cw),
    Modularity_FG = modularity(cfg),
    Num_Communities_Walk = length(unique(membership(cw))),
    Num_Communities_FG = length(unique(membership(cfg))),
    Diameter = diameter(g) 
  )
  results_hub <- rbind(results_hub, row)
}

print(results_hub)
```




```{r community-detection-visualization, cache = TRUE}
all_colors <- colors()
vibrant_colors <- all_colors[!grepl("gray|grey", all_colors)]
unique_palette <- sample(vibrant_colors)

# Loop through each subgraph in your list
for(i in seq_along(subgraph_list_hub)) {
  g <- subgraph_list_hub[[i]]
  
  cw  <- cluster_walktrap(g)
  cfg <- cluster_fast_greedy(g)
  
  # Walktrap
  num_comms_cw <- length(unique(membership(cw)))
  v_colors_cw <- unique_palette[1:num_comms_cw]
  
  plot(cw, g,
       vertex.size = 4,
       vertex.label = NA,
       vertex.color = v_colors_cw[membership(cw)],
       vertex.frame.color = "gray40",
       edge.color = "black",
       edge.width = 0.5,
       edge.curved = 0.3,
       col = "grey95",      # hull color
       border = "red",      # hull border
       main = paste("Hub", i, ": Walktrap"))

  # Fast & Greedy
  num_comms_cfg <- length(unique(membership(cfg)))
  v_colors_cfg <- unique_palette[1:num_comms_cfg]
  
  plot(cfg, g,
       vertex.size = 4,
       vertex.label = NA,
       vertex.color = v_colors_cfg[membership(cfg)],
       vertex.frame.color = "gray40",
       edge.color = "black",
       edge.width = 0.5,
       edge.curved = 0.3,
       col = "grey95",
       border = "blue",     # changed border color to distinguish
       main = paste("Hub", i, ": F&G"))
}
```



```{r adj-mat-reorder, cache = TRUE}
# Original (unordered) adjacency matrix
A <- as_adjacency_matrix(sub1, sparse = TRUE)
image(A, main = "Adjacency Matrix - original node order", col.region = c("red"),
      xlab = "Node", ylab = "Node")

# Reorder by Walktrap community membership
ord <- order(cw$membership)
A_reordered <- A[ord, ord]

image(A_reordered,
      main = "Adjacency Matrix — reordered by community (Walktrap)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")

# Reorder by Fast Greedy community membership
ord <- order(cfg$membership)
A_reordered1 <- A[ord, ord]

image(A_reordered1,
      main = "Adjacency Matrix — reordered by community (Fast Greedy)", col.region = c("red"),
      xlab = "Node (reordered)", ylab = "Node (reordered)")
```
       

